# COM-Poisson models with varying dispersion # {#chapter3-cmpvarying}

```{block2, echo=TRUE, type="abstract_en"}

There are many experiments that experimental conditions can have an
effect on mean and dispersion. In this case, models with constant
dispersion can produce a loss of efficiency. In this chapter, we propose
an extension of the COM-Poisson model to jointly model the mean and the
dispersion. This approach allows analyzing the data that exhibit under-
and overdispersion depending on covariates. The estimation and inference
are based on likelihood method. We carried out simulation studies to
verify the finite sample properties of the maximum likelihood
estimators. The results from our simulation study show that the maximum
likelihood estimators are unbiased and consistent for both mean and
dispersion regression parameters. The application of the COM-Poisson
models with varying dispersion is illustrated with the analysis of two
experimental datasets. The computational routines for fitting the models
and the datasets are available in the appendix.
\vspace*{0.3cm}

**Keywords:** COM-Poisson distribution, Double generalized linear
  models, GAMLSS, Varying dispersion.

```

## Introduction ## {#chap3-introduction}

The standard Gaussian linear models are based on the assumption of
variance homogeneity [@Aitkin1987]. Generalized linear models relax this
assumption by assuming the observations come from some distribution in
the exponential family [@Nelder1972]. A key feature of exponential
family distribution is the so-called mean-variance relationship,
i.e. the variance is a deterministic function of the mean $\text{Var}(Y)
= \phi V(\mu)$. The main examples are $V(\mu) = \mu(1-\mu)$ for the
binomial distribution, $V(\mu) = \mu$ for the Poisson distribution,
$V(\mu) = \mu^2$ for the gamma distribution and $V(\mu) = \mu^3$ for the
inverse-Gaussian distribution [@McCullagh1989]. This allows modeling
specific heterogeneity by assuming an appropriate distribution. However,
once the mean-variance relationship is specified, the variance is
assumed to be known up to a constant of proportionality, the dispersion
parameter~$\phi$. To ensure more flexibility in the analysis of
non-Gaussian heterogeneous data, we explore methods to model dispersion
depending on covariates. We focus on count data only.

Modeling dispersion depending on covariates in the analysis of count
data has not been much explored in the literature. The class of double
generalized linear models [@Smyth1988; @McCullagh1989; @Smyth1999] can
be used to do it. This class was widely explored for continuous
data. @Smyth1988 discussed this class using a double generalized linear
model based on the gamma distribution. Adjusted likelihood methods for
estimation and inference are presented by @Smyth1999. @Paula2013b
discussed diagnostics for this class and used a assumed gamma
distributed data as an application. @Andersen2017 extended the double
generalized linear model by considering compound Poisson
distributions. Related to discrete data, @Vieira2011 proposed a Bayesian
analysis of the double generalized linear models for binomial data.

Another approach that has gained momentum in the last decade is the
generalized additive models for location, shape, and scale (GAMLSS)
[@Rigby2005]. This approach extends the generalized linear models in
different directions: (i) the distribution for the response variable can
be selected from a more general family; the only restriction is that the
derivatives with respect to each parameter must be computable, (ii) all
parameters of the distribution can be depended on covariates, and (iii)
the systematic relationship between covariates and parameters can be
parametric (linear predictor) or nonparametric (smooth) functions.

In this chapter, we propose to jointly model the mean and dispersion
based on the COM-Poisson distribution. This approach is very similar to
the GAMLSS, however, we develop and explore our own estimation
methods. This approach allows data exhibit under- and overdispersion
depending on experimental conditions.

This chapter is organized as follows. In Section \@ref(chap3-dglm), we
briefly describe and discuss the double generalized linear models. The
newly proposed COM-Poisson models is considered in Section
\@ref(chap3-cmpreg). In Section \@ref(chap3-estimation-and-inference),
we present estimation and inference for the COM-Poisson regression model
in a likelihood framework. Section
\@ref(chap3-applications-and-discussion) is devoted to illustrate the
application of the COM-Poisson model with varying dispersion for the
analysis of two data sets. We compare the results of COM-Poisson model
with double generalized linear models and GAMLSS approach. Finally,
concluding remarks close the chapter in Section
\@ref(chap3-final-remarks). We provide an `R` implementation for fitting
the COM-Poisson models with varying dispersion, together with the
analyzed data sets, ins the package `cmpreg`. Illustrative codes are
presented in the appendix.

## Double generalized linear models ## {#chap3-dglm}

Following @Smyth1988 and McCullagh and Nelder
[-@McCullagh1989, chap. 10], the so-called double generalized linear
models (DGLM) can be used to model mean and dispersion jointly by
considering two linked generalized linear models (GLM). Let $y_i$,
$i=1,2,\ldots,n$ be a set of independent observations of $Y_i$ and
$\bm{x}_i = (x_{i1}, x_{i2}, \ldots, x_{np})^\top$ a vector of known
covariates. The standard GLMs assume that
$$
\text{E}(Y_i) = \mu_i, \quad
g(\mu_i) = \eta_i = \bm{x}_i^\top \bm{\beta}
\quad\text{and}\quad
\text{Var}(Y_i)=\phi V(\mu_i),
$$
where $g(.)$ is a suitable link function and $V(.)$ is the variance
function. The DGLMs generalize the GLMs to allow the dispersion
parameter varying across observations, $\text{Var}(Y_i)=\phi_i
V(\mu_i)$. The model for dispersion is specified by
$$
\text{E}(d_i) = \phi_i, \quad
h(\phi_i) = \xi_i = \bm{z}_i^\top \bm{\gamma}
\quad\text{and}\quad
\text{Var}(d_i)=\tau V_d(\phi_i),
$$
where $d_i$ is a suitable statistic chosen as a measure of dispersion,
$h(.)$ is the dispersion link function and $V_d(.)$ is the dispersion
variance function. The commom choice for $d_i$ is the deviance
components from the mean model. Based on asymptotic distribution
of $d_i$, a gamma model with $V_d(\phi_i) = 2\phi_i^2$ is a suitable
natural choice for the dispersion model.

The parameter estimation proposed is based on a two-step
iterative algorithm: (i) holding $\bm{\gamma}$ fixed we obtain the
dispersion components and estimate $\bm{\beta}$; and (ii) fixing the
estimated value of $\bm{\beta}$, we obtain the deviance components and
estimate $\bm{\gamma}$. These two steps are then alternated until
convergence.

The idea of using a interlinked pair of generalized linear models for
allowing mean and dispersion effects depend on covariates was first put
forward by @Pregibon1984 and further developed by different authors with
different names. @Smyth1988 called it generalized linear models with
varying dispersion, while @McCullagh1989, at the same time, named this
approach by joint modelling of mean and dispersion. Nowadays, the most
used name to refer to this approach is double generalized linear models,
introduced by @Smyth1999.

Another approach that has become popular for modeling mean and
dispersion is the class of generalized additive models for location,
scale, and shape (GAMLSS) [@Rigby2005]. It is important to note that the
DGLMs for discrete data are not a particular case of GAMLSS. Although
both can model mean and dispersion jointly in a similar way, the GAMLSS
require a two-parameter distribution for the response. In this chapter,
we compare the proposed COM-Poisson model varying dispersion with a
GAMLSS based on the double Poisson distribution [@Efron1986; @Zou2013]
and a DGLM Poisson.

## Modeling mean and dispersion COM-Poisson ## {#chap3-cmpreg}

The COM-Poisson distribution is a two-parameter generalization of the
Poisson distribution that can deal with under-, over- and equidispersion
[@Shmueli2005; @Sellers2010]. The probability mass function of the
COM-Poisson distribution is given by
\begin{equation}
  \label{eqn:chap3-pmf-cmp0}
  \Pr(Y = y) = \frac{\lambda^y}{(y!)^\nu \text{Z}(\lambda, \nu)},
  \quad \text{where} \quad
  \text{Z}(\lambda,\nu) = \sum_{j=0}^\infty \frac{\lambda^j}{(j!)^\nu}.
\end{equation}
The series $\text{Z}(\lambda, \nu)$ is a normalizing constant that
cannot be expressed in closed form unless for special cases. The
parameters $\lambda$ and $\nu$ can be seen in terms of the ratios of
successive probabilities, $\Pr(Y = y - 1)/\Pr(Y = y) = y^\nu/\lambda$.

The parameter $\nu$ is the dispersion parameter with a clear
interpretation; overdispersion for $0 < \nu < 1$ and underdispersion
for $\nu>1$.  When $\nu=1$, the Poisson distribution results as a
special case. On the other hand, the parameter $\lambda$ has no clear
interpretation, unless for $\nu=1$, and it is strongly related to
$\nu$. To circumvent this problem, @RibeiroJr2018 propose a
re-parameterization of the COM-Poisson distribution to provide an
approximate mean parameter. By introducing the new parameter $\mu>0$,
$$
\mu = \lambda^{1/\nu} - \frac{\nu - 1}{2\nu}
  \quad \Rightarrow \quad
  \lambda = \left (\mu +
    \frac{(\nu - 1)}{2\nu} \right )^\nu,
$$
the authors showed that the new parametrization has good properties for
estimation and inference and proposed a regression model on approximated
mean. In this paper, we propose to allow both $\mu$ and $\nu$
parameters depending on covariates.

Let $y_i$, $i=1,2,\ldots,n$ be a set of independents realizations of
$Y_i$ following a COM-Poisson distribution with parameters $\mu_i$ and
$\nu_i$. The proposed COM-Poisson varying dispersion model assumes
$$
\eta_i = g(\mu_i) = \bm{x}_i^\top \bm{\beta} \quad\text{and}\quad
\xi_i = h(\nu_i) = \bm{z}_i^\top \bm{\gamma},
$$
where $\bm{\beta} = (\beta_1, \beta_2, \ldots, \beta_p)^\top$ and
$\bm{\gamma} = (\gamma_1, \gamma_2, \ldots, \gamma_q)^\top$ are the
parameters to be estimated, $\bm{x}_i = (x_{i1}, x_{i2}, \ldots,
x_{np})^\top$ and $\bm{z}_i = (z_{i1}, z_{i2}, \ldots, z_{nq})^\top$ are
vectors of known covariates, and $g(.)$ and $h(.)$ are suitable link
functions. We use the logarithmic link function for both mean and
dispersion.

The COM-Poisson model with varying dispersion is quite similar to the
double generalized linear approach, both extend the generalized linear
models to model the dispersion as well as the mean. However, COM-Poisson
with varying dispersion is a fully parametric model that has the
advantages of allowing predictions for probabilities and generalizations
such as random effects and modeling of censored data.

## Estimation and inference ## {#chap3-estimation-and-inference}

To fit COM-Poisson models with varying dispersion, we use the maximum
likelihood estimation method. The log-likelihood function for
$\bm{\theta} = (\bm{\beta}^\top, \bm{\gamma}^\top)^\top$ parameters
is given by
\begin{equation}
\label{eqn:ll-cmpvarying}
  \ell = \ell(\bm{\theta} ; \bm{y}) = \sum_{i=1}^{n} \left \{
    \nu_i\log \left ( \mu_i + \frac{\nu_i-1}{2\nu_i}\right ) -
    \nu_i \log(y_i) - \log[Z(\mu_i, \nu_i)]
  \right \},
\end{equation}
where $\mu_i = \exp(\bm{x}_i^\top \bm{\beta})$, $\nu_i =
\exp(\bm{z}_i^\top \bm{\gamma})$, and $Z(\mu_i, \nu_i)$ is a normalizing
constant computed for the parameters $\mu_i$ and $\nu_i$.

Parameter estimation requires the numerical maximization of Equation
(\@ref(eqn:ll-cmpvarying)). Since the derivatives of $\ell$ cannot be
obtained in closed forms, we compute them by central finite differences
using the Richardson method as implemented in package `numDeriv`
[@numDeriv-pkg] for the statistical software `R` [@Rcore2018]. Based on
the orthogonality property between $\bm{\mu}_i$ and $\bm{\nu}_i$
[see @RibeiroJr2018], we set two strategies to obtain the maximum
likelihood estimates:
\vspace{-0.2cm}
\begin{itemize}
  \setlength\itemsep{0em}
  \item[(a)] the \textit{joint} strategy, where we obtain
  $\hat{\bm{\beta}}$ and $\hat{\bm{\gamma}}$ jointly by maximizing
  $\ell(\bm{\theta} ; \bm{y})$; and
  \item[(b)] the \textit{two-stages} strategy, where we fix
  $\hat{\bm{\beta}}$ at the maximum likelihood estimates of Poisson
  model and obtain $\hat{\bm{\gamma}}$ by maximizing $\ell(\bm{\gamma};
  \bm{\beta} = \hat{\bm{\beta}}, \bm{y})$ and then estimate the Hessian
  matrix at $(\hat{\bm{\beta}}, \hat{\bm{\gamma}})$.
\end{itemize}

Although we have an additional step for estimating the Hessian matrix,
the strategy (b) tends to be faster than (a), because of the
maximization is performed in $\mathbb{R}^q$ whereas, for strategy (a),
the estimation is an optimization problem in $\mathbb{R}^{p+q}$ space.

Standard errors for the parameter estimates are obtained based on the
observed information matrix. Let the derivatives be computed at the
maximum likelihood estimates. The variance and covariance matrix of the
maximum likelihood estimators may be expressed as
$$
\bm{V}_\theta=
\begin{pmatrix}
    -\partial \ell^2/
        \partial\bm{\beta}\partial\bm{\beta}^\top &
    -\partial \ell^2/
        \partial\bm{\beta}\partial\bm{\gamma}^\top \\
    -\partial \ell^2/
        \partial\bm{\gamma}\partial\bm{\beta}^\top &
    -\partial \ell^2/
        \partial\bm{\gamma}\partial\bm{\gamma}^\top
\end{pmatrix}^{-1} =
\begin{pmatrix}
    \bm{V}_{\beta\phantom{\gamma}} & \bm{V}_{\beta\gamma} \\
    \bm{V}_{\gamma\beta} & \bm{V}_{\gamma\phantom{\beta}} \\
\end{pmatrix}.
$$

Based on the asymptotic distribution of the maximum likelihood
estimators, we obtain the variances for $\hat{\eta}_i$ and $\hat{\xi}_i$
using delta method,
$\text{Var}(\hat{\eta}_i) = \bm{x}_i^\top \bm{V}_{\beta|\gamma}\bm{x}_i$
and
$\text{Var}(\hat{\xi}_i) = \bm{z}_i^\top \bm{V}_{\gamma|\beta}\bm{z}_i$,
where
$\bm{V}_{\beta|\gamma} = \bm{V}_{\beta} - \bm{V}_{\beta\gamma}
\bm{V}_{\gamma}^{-1} \bm{V}_{\gamma\beta}$
and
$\bm{V}_{\gamma|\beta} = \bm{V}_{\gamma} - \bm{V}_{\gamma\beta}
  \bm{V}_{\beta}^{-1} \bm{V}_{\beta\gamma}$.
Since $\bm{\beta}$ and $\bm{\gamma}$ were found to be nearly orthogonal,
$\bm{V}_{\beta\gamma} = \bm{V}_{\gamma\beta}^\top \approx \bm{0}$, hence
$\bm{V}_{\beta|\gamma} \approx \bm{V}_{\beta}$ and
$\bm{V}_{\gamma|\beta} \approx \bm{V}_{\gamma}$, in other words, inference
based on the conditional log-likelihood and inference based on the marginal
log-likelihood is the same.

Confidence intervals for $\mu_i$ and $\nu_i$ are obtained by
transforming the confidence intervals for $\eta_i$ and $\xi_i$.  We
implemented the two strategies for fitting the COM-Poisson models and
the methods for computing the confidence intervals in the
`cmpreg`^[Available on GitHub https://github.com/jreduardo/cmpreg]
package for the software `R`.

## Simulation study ## {#chap3-simulation-study}

In this section, we present a simulation study designed to assess the
properties of the maximum likelihood estimators using the proposed (a)
joint and (b) two-stages strategies. We compared the estimation
strategies by maximized log-likelihoods and fitting times.

We considered average counts varying from 3 to 30 and average dispersion
varying from 0.3 to 4 arising from regression models with a continuous
($x_1$) and a categorical ($x_2$) covariate. The continuous covariate
was generated as a linearly increasing sequence from 0 to 1 with length
equal to the sample size. Similarly, the categorical covariate was
generated as a sequence of three values each one repeated n/3 times
(rounding up when required), where $n$ denotes the sample size. The
parameter $\mu$ and $\nu$ of the reparametrized COM-Poisson random
variable is given by $\log(\mu) = \beta_0 + \beta_1 x_1 + \beta_{21}
x_{21} + \beta_{22} x_{22}$ and $\log(\nu) = \gamma_0 + \gamma_1 x_1 +
\gamma_{21} x_{21} + \gamma_{22} x_{22}$, respectively, where $x_{21}$
and $x_{22}$ are dummy representing the levels of $x_2$. The regression
coefficients were fixed at the values, $\beta_0 = 2, \beta_1 = 0.5,
\beta_{21} = 0.8$ and $\beta_{22}=-0.8$ for the mean, and $\gamma_0 =
-0.7, \gamma_1 = 1.3, \gamma_{21} = 0.8$ and $\gamma_{22}=-0.6$ for the
dispersion. Figure \@ref(fig:chap3-justpars) shows the parameter space
evaluated by this simulation study. With this simulation design, we are
able to assess the properties of the parameter estimators in situation
of high and low counts and equi-, under-, and overdispersion.

```{r chap3-justpars, results="asis", fig.width=8.5, fig.height=4, fig.cap="Values for the mean (left) and for the dispersion (right) according to the regression models adopted."}
knitr::include_graphics("chapter3-figure/justpars-1")
```

In order to check the consistency of the estimators we considered four
different sample sizes: $50, 100, 300$ and $1000$; generating $1000$
data sets in each case. In Figure \@ref(fig:chap3-bias-plot), we show
the bias of the estimators along with the confidence intervals
calculated as average bias plus and minus $1.96$ times the average
standard error. The scales are standardized for each parameter by
dividing the average bias by the average standard error obtained for the
sample of size $50$. The results show that for both estimation methods
the expected bias and the standard error tend to 0 as the sample size
increases. Thus, this shows the unbiasedness, consistency, and symmetry
of the empirical distribution of the maximum likelihood estimators using
both strategies.

```{r chap3-bias-plot, results="asis", fig.width=9, fig.height=4, fig.cap="Distributions of standardized bias (gray box-plots) and average with confidence intervals (black segments) by different sample sizes."}
knitr::include_graphics("chapter3-figure/bias-plot-1")
```

In order to compare the proposed strategies, we compute the
log-likelihood function at the maximum likelihood estimates for each
simulated data set. The results are shown in Figure
\@ref(fig:chap3-loglik-plot). The maximized log-likelihood obtained
using the strategy (b) is always smaller than that obtained considering
the strategy (a). This is expected since in strategy (b) we do not refit
the mean regression parameters. However, the likelihoods values are
similar between the two strategies, showing that the estimates of the
Poisson model are very close to the genuine mean regression maximum
likelihood estimates of the COM-Poisson.

```{r chap3-loglik-plot, results="asis", fig.width=9, fig.height=4, fig.cap="Scatter plots of maximized log-likelihood obtained using the strategy (a) $vs$. those obtained using the strategy (b)."}
knitr::include_graphics("chapter3-figure/loglik-plot-1")
```

The computational times to fit the models using the both strategies are
shown in Table \@ref(tab:chap3-timestofit). As highlighted in the
previous section, we expect the strategy (b) to be faster, since the
maximization is done in a smaller dimension. The results show that the
estimation by using the strategy (a) takes, on average, around 30\% more
time than using the strategy (b).

\begin{table}[ht]
\centering
\begingroup
\caption{Average and quartiles of the time to fit the COM-Poisson models
using strategies (a) and (b), in seconds, and ratios between the averages.}
\label{tab:chap3-timestofit}
\begin{tabular*}{\textwidth}{L{1.5cm}rrrrrrr}
\toprule
& \multicolumn{3}{c}{Strategy (a)} &
\multicolumn{3}{c}{Strategy (b)} & \\
\cmidrule(lr){2-4} \cmidrule(lr){5-7}
Sample size & Average & 1st quartile & 3rd quartile & Average & 1st
 quartile & 3rd quartile &  Ratio\\
  \midrule
   50 & 1.01 & 0.98 & 1.07 & 0.77 & 0.77 & 0.82 & 1.31 \\
   100 & 1.80 & 1.64 & 1.94 & 1.37 & 1.26 & 1.50 & 1.31 \\
   300 & 3.89 & 3.76 & 4.78 & 2.98 & 2.92 & 3.60 & 1.30 \\
   1000 & 14.24 & 12.37 & 15.87 & 11.13 & 9.70 & 12.47 & 1.28 \\
   \bottomrule
\end{tabular*}
\endgroup
\end{table}

## Applications and discussion ## {#chap3-applications-and-discussion}

In this section, we shall present two data analyzes to illustrate the
application of the COM-Poisson model with varying dispersion. We compare
the results of the COM-Poisson model with the double generalized linear
models approach, fitted using `dglm::dglm(..., family = poisson)`, and
double Poisson model, fitted using `gamlss::gamlss(..., family =
DPO)`. The data sets and `R` codes used in this section are available in
the appendix.

### Analysis of nitrofen experiment ###

This dataset comes from a completely randomized experiment where the
number of live offspring of a species of zooplankton were recorded for
different doses of nitrofen. These data are presented in Section
\@ref(chap0-nitrofen) and were analyzed by @RibeiroJr2018 using the
COM-Poisson model with constant dispersion.

Here, to analyze the number of live offspring under different doses of
nitrofen, we consider the following linear predictors for the dispersion
\vspace{0.2cm}
\begin{tabular}{ll}
    Constant:  & $\log(\nu_{ij}) = \gamma_0$,\\
    Linear:    & $\log(\nu_{ij}) = \gamma_0 +
                  \gamma_1 \texttt{x}_{1i}$,\\
    Quadratic: & $\log(\nu_{ij}) = \gamma_0 +
                  \gamma_1 \texttt{x}_{1i} +
                  \gamma_2 \texttt{x}_{2i}^2$ e\\
    Cubic:     & $\log(\nu_{ij}) = \gamma_0 +
                  \gamma_1 \texttt{x}_{1i} +
                  \gamma_2 \texttt{x}_{2i}^2 +
                  \gamma_3 \texttt{x}_{3i}^3$,
  \end{tabular}
\vspace{0.2cm}
\noindent
where $i$ and $j$ refers to the nitrofen concentration level (`dose`)
and to the replicates, respectively. The covariates $\bm{\texttt{x}}_1$,
$\bm{\texttt{x}}_2$ and $\bm{\texttt{x}}_3$ are the orthogonal linear,
quadratic and cubic polynomials of the `dose`. To the mean, we consider
a raw cubic polynomial of `dose`.

In Table \@ref(tab:chap3-anova), some goodness-of-fit measures (minus
two times the maximized log-likelihood and Akaike information criterion)
are presented together with corresponding likelihood ratio tests between
nested models. The results show a clear evidence of that the dispersion
of the counts is at least linearly dependent on the nitrofen
concentration level. Consider the COM-Poisson model, the difference of
deviances of the models considering the quadratic and linear predictors
is greater than the difference obtained for double Poisson and double
GLM Poisson models.

\begin{table}[ht]
  \centering
  \caption{Nitrofen data: goodness-of-fit measures (deviance and AIC) and model comparisons.}
  \label{tab:chap3-anova}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcccrl}
    \toprule
 &     & \multicolumn{4}{c}{COM-Poisson} \\
               \cmidrule(lr){3-6}
 & G.l & Deviance & AIC & $\rchi^2$ & $\Pr(>\rchi^2$) \\
  \midrule
  Constant & 45 & 288.127 & 298.127 &  &  \\
  Linear & 44 & 274.111 & 286.111 & 14.0164 & 0.0002 \\
  Quadratic & 43 & 270.493 & 284.493 & 3.6179 & 0.0572 \\
  Cubic & 42 & 269.503 & 285.503 & 0.9898 & 0.3198 \\
   \bottomrule
  \end{tabular*}\\[0.15cm]
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcccrl}
 &     & \multicolumn{4}{c}{double Poisson (\texttt{GAMLSS})} \\
               \cmidrule(lr){3-6}
 & G.l & Deviance & AIC & $\rchi^2$ & $\Pr(>\rchi^2$) \\
  \midrule
  Constant & 45 & 288.181 & 298.181 &  &  \\
  Linear & 44 & 273.530 & 285.530 & 14.6512 & 0.0001 \\
  Quadratic & 43 & 271.204 & 285.204 & 2.3256 & 0.1273 \\
  Cubic & 42 & 269.201 & 285.201 & 2.0031 & 0.1570 \\
   \bottomrule
  \end{tabular*}\\[0.15cm]
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcccrl}
 &     & \multicolumn{4}{c}{double GLM Poisson (\texttt{DGLM})} \\
               \cmidrule(lr){3-6}
 & G.l & Deviance & AIC & $\rchi^2$ & $\Pr(>\rchi^2$) \\
  \midrule
  Constant & 45 & 287.690 & 297.690 &  &  \\
  Linear & 44 & 272.634 & 284.634 & 15.0565 & 0.0001 \\
  Quadratic & 43 & 269.997 & 283.997 & 2.6368 & 0.1044 \\
  Cubic & 42 & 268.165 & 284.165 & 1.8317 & 0.1759 \\
   \bottomrule
   \end{tabular*}
   \footnotesize \raggedright
   Deviance is computed as minus twice log-likelihood
\end{table}

\begin{table}[ht]
\centering
\caption{Nitrofen data: Parameter estimates (Est) and standard errors (SEs) for the fitted double regression COM-Poisson model.}
\label{tab:chap3-coef-table-nitrofen}
\begingroup
\begin{tabularx}{\textwidth}{m{.8cm}lR{3.1cm}R{3.1cm}R{3.1cm}R{3.1cm}}
  \toprule
           & & \multicolumn{4}{c}{Estimates (Standard Errors)} \\
               \cmidrule(lr){3-6}
 Parameter & & Constant & Linear & Quadratic & Cubic \\
  \midrule
 Mean &  &  &  &  & \\
   & $\beta_0$ & $3.4769$ $(0.0541)^\text{a}$ & $3.4778$ $(0.0283)^\text{a}$ & $3.4777$ $(0.0339)^\text{a}$ & $3.4776$ $(0.0360)^\text{a}$ \\
   & $\beta_1$ & $-0.0879$ $(0.1943)^{\phantom{\text{a}}}$ & $-0.1055$ $(0.1424)^{\phantom{\text{a}}}$ & $-0.1167$ $(0.1322)^{\phantom{\text{a}}}$ & $-0.1153$ $(0.1283)^{\phantom{\text{a}}}$ \\
   & $\beta_2$ & $0.1547$ $(0.1731)^{\phantom{\text{a}}}$ & $0.1747$ $(0.1492)^{\phantom{\text{a}}}$ & $0.1917$ $(0.1375)^{\phantom{\text{a}}}$ & $0.1886$ $(0.1318)^{\phantom{\text{a}}}$ \\
   & $\beta_3$ & $-0.0976$ $(0.0396)^\text{a}$ & $-0.1028$ $(0.0380)^\text{a}$ & $-0.1082$ $(0.0368)^\text{a}$ & $-0.1069$ $(0.0350)^\text{a}$ \\
 Dispersion &  &  &  &  & \\
   & $\gamma_0$ & $0.0474$ $(0.2047)^{\phantom{\text{a}}}$ & $0.2948$ $(0.2112)^{\phantom{\text{a}}}$ & $0.2437$ $(0.2589)^{\phantom{\text{a}}}$ & $0.3532$ $(0.2268)^{\phantom{\text{a}}}$ \\
   & $\gamma_1$ & -- & $-5.2441$ $(1.3630)^\text{a}$ & $-7.0024$ $(2.3054)^\text{a}$ & $-5.7301$ $(1.8440)^\text{a}$ \\
   & $\gamma_2$ & -- & -- & $-3.9807$ $(2.4439)^{\phantom{\text{a}}}$ & $-2.9175$ $(1.9045)^{\phantom{\text{a}}}$ \\
   & $\gamma_3$ & -- & -- & -- & $1.5221$ $(1.4119)^{\phantom{\text{a}}}$ \\
   \bottomrule
\end{tabularx}
\endgroup
\footnotesize \raggedright
 Est (SE)$^\text{a}$ indicates $|$Est$/$SE$|$ $> 1,96$.
\end{table}

The estimated parameters, their associated standard errors and
individual Wald tests are presented in Table
\@ref(tab:chap3-coef-table-nitrofen) for COM-Poisson model. We fitted
the models using the strategy (a), for that reason the parameter
estimates for mean structure are slightly different for the different
predictors. For the dispersion structure, there is no evidence to keep
the quadratic term, the Wald's statistics is equals to $-1.63$
($p$-value $=0.1034$).

Figure \@ref(fig:fitted-curves-nitrofen) presents the fitted mean values
$\hat{\mu}_i$ and fitted dispersion values $\hat{\nu}_i$ across doses
between 0 and 3.1$\mu$g$/10^2$litre for the four different linear
predictors adopted to dispersion. When we consider the dispersion
constant, the model indicates equidispersion ($\nu=1$). However, when we
relax this assumption, it is clear that the dispersion change across
nitrofen levels. In particular, all models show that around
2$\mu$g$/10^2$litre the number of live offspring change from under- to
overdispersed.

```{r fitted-curves-nitrofen, results="asis", fig.width=9, fig.height=5, fig.cap="Fitted mean and dispersion values with 95\\% confidence intervals for the considered linear predictors."}
knitr::include_graphics("chapter3-figure/fitted-curves-nitrofen-1")
```

```{r fitted-moments-nitrofen, results="asis", out.width="75%", fig.width=7, fig.height=4, fig.cap="Expected values and variances obtained from the fitted models with linear and quadratic predictors for the dispersion. The points are the sample means and sample variances, respectively."}
knitr::include_graphics("chapter3-figure/fitted-moments-nitrofen-1")
```

The expected values and variances for nitrofen doses between 0 and
3.1$\mu$g$/10^2$litre, obtained from the fitted models with linear and
quadratic predictors for the dispersion, are presented in Figure
\@ref(fig:fitted-moments-nitrofen). Note that although the polynomial
adopted for the mean is not the saturated one, the fitted curve almost
interpolates the sample averages. The predicted variances show
interesting behavior. Note that these curves are nonlinear functions of
$\mu_i$ and $\nu_i$, so a linear predictor of $\nu_i$ does not imply a
linear curve here. The model with a linear predictor for the dispersion
does not fit the sample variances for the doses 1.6 and
2.35$\mu$g$/10^2$litre well. The model with a quadratic predictor fits
these variances better, however, it produces rather strange behavior for
values greater than 2.35$\mu$g$/10^2$litre.

### Analysis of soybean experiment ###

In this second application, we analyze the data resulting from a
$5\times3$ factorial experiment conducted in a randomized complete block
design, see Section \@ref(chap0-soy). In this experiment, we have two
counting responses recorded, number of grains and number of pods. Here
we focus on analyzing the number of pods and consider the following
predictors for mean and dispersion
$$
\begin{aligned}
  \log(\mu_{ijk}) &= \beta_0 + \kappa_i + \tau_j +
    (\beta_1 + \delta_j)\texttt{K}_k + \beta_2^2 \texttt{K}_k^2\\
  \log(\nu_{ijk}) &= \alpha_j + \gamma_1\texttt{K}_k,
\end{aligned}
$$
where $i$, $j$, and $k$ varies according to block, moisture level, and
potassium fertilization. Note that for the dispersion, the $\alpha_j$ is
the logarithm of dispersion parameter when the potassium dose is 0
(without an intercept term).

In Table \@ref(tab:chap3-anova-soybean), we present the goodness-of-fit
measures (minus twice log-likelihood and Akaike information criterion)
together with the corresponding likelihood ratio tests between the
current model (moisture + K) and the reduced models for dispersion:
$\gamma_1=0$ (moisture effect) and $\alpha_j = \alpha$ (constant). The
log-likelihood achieved for the COM-Poisson, double Poisson and double
generalized linear Poisson models are practically the same for all three
predictors for dispersion. The results show a significant improvement in
varying the dispersion for each moisture level and do not present a
clear relation between the dispersion and the potassium doses when
moisture is already included in the linear predictor.

\begin{table}[ht]
  \centering
  \caption{Soybean data (pods): goodness-of-fit measures (deviance and AIC) and model comparisons.}
  \label{tab:chap3-anova-soybean}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcccrl}
    \toprule
 &     & \multicolumn{4}{c}{COM-Poisson} \\
               \cmidrule(lr){3-6}
 & Df & Deviance & AIC & $\rchi^2$ & $\Pr(>\rchi^2$) \\
  \midrule
  Constant & 62 & 534.232 & 558.232 &  &  \\
  Moisture & 60 & 527.866 & 555.866 & 6.3658 & 0.0415 \\
  Moisture + K & 59 & 525.848 & 555.848 & 2.0181 & 0.1554 \\
   \bottomrule
  \end{tabular*}\\[0.15cm]
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcccrl}
 &     & \multicolumn{4}{c}{double Poisson (\texttt{GAMLSS})} \\
               \cmidrule(lr){3-6}
 & Df & Deviance & AIC & $\rchi^2$ & $\Pr(>\rchi^2$) \\
  \midrule
  Constant & 62 & 534.231 & 558.231 &  &  \\
  Moisture & 60 & 527.875 & 555.875 & 6.3566 & 0.0417 \\
  Moisture + K & 59 & 525.874 & 555.874 & 2.0005 & 0.1572 \\
   \bottomrule
  \end{tabular*}\\[0.15cm]
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcccrl}
 &     & \multicolumn{4}{c}{double GLM Poisson (\texttt{DGLM})} \\
               \cmidrule(lr){3-6}
 & Df & Deviance & AIC & $\rchi^2$ & $\Pr(>\rchi^2$) \\
  \midrule
  Constant & 62 & 534.042 & 558.042 &  &  \\
  Moisture & 60 & 527.677 & 555.677 & 6.3654 & 0.0415 \\
  Moisture + K & 59 & 525.663 & 555.663 & 2.0137 & 0.1559 \\
   \bottomrule
   \end{tabular*}
   \footnotesize \raggedright
   Deviance is computed as $-2\times$logLik.
\end{table}

Table \@ref(tab:chap3-coef-table-soybean) gives the parameter estimates
and associated standard errors for the three different predictors. It
seems that the mean structure is well specified since the Wald tests
lead to the same conclusions for the different dispersion
structures. The likelihood ratio tests indicate that we have different
dispersion levels for each moisture level. In particular, considering
the reduced model $\gamma_1=0$, the fitted dispersions values for each
moisture level are $\hat{\nu}_{\text{37.5\%}}=0.701$,
$\hat{\nu}_{\text{50.0\%}}=0.742$, and
$\hat{\nu}_{\text{63.5\%}}=1.99$. The point estimates indicate
overdispersion for the two first and underdispersion for the latter.

\begin{table}[ht]
\centering
\caption{Soybean data (pods): Parameter estimates (Est) and standard errors (SEs) for the fitted double regression COM-Poisson model.}
\label{tab:chap3-coef-table-soybean}
\begingroup
\begin{tabularx}{\textwidth}{m{1.3cm}lR{3.8cm}R{3.8cm}R{3.8cm}}
  \toprule
           & & \multicolumn{3}{c}{Estimates (Standard Errors)} \\
               \cmidrule(lr){3-5}
 Parameter &  & Constant & Moisture & Moisture + K \\
  \midrule
 Mean &  &  &  & \\
   & $\beta_0$ & $4.0062$ $(0.0544)^\text{a}$ & $4.0112$ $(0.0576)^\text{a}$ & $4.0273$ $(0.0632)^\text{a}$ \\
   & $\kappa_{\text{II}}$ & $-0.0293$ $(0.0426)^{\phantom{\text{a}}}$ & $-0.0320$ $(0.0381)^{\phantom{\text{a}}}$ & $-0.0398$ $(0.0359)^{\phantom{\text{a}}}$ \\
   & $\kappa_{\text{III}}$ & $-0.0727$ $(0.0431)^{\phantom{\text{a}}}$ & $-0.0991$ $(0.0393)^\text{a}$ & $-0.1097$ $(0.0379)^\text{a}$ \\
   & $\kappa_{\text{IV}}$ & $-0.1254$ $(0.0437)^\text{a}$ & $-0.1452$ $(0.0403)^\text{a}$ & $-0.1573$ $(0.0380)^\text{a}$ \\
   & $\kappa_{\text{V}}$ & $-0.1037$ $(0.0446)^\text{a}$ & $-0.0960$ $(0.0411)^\text{a}$ & $-0.1016$ $(0.0394)^\text{a}$ \\
   & $\tau_{\text{50.0\%}}$ & $0.1573$ $(0.0583)^\text{a}$ & $0.1571$ $(0.0659)^\text{a}$ & $0.1557$ $(0.0723)^\text{a}$ \\
   & $\tau_{\text{63.5\%}}$ & $0.1730$ $(0.0582)^\text{a}$ & $0.1731$ $(0.0562)^\text{a}$ & $0.1758$ $(0.0610)^\text{a}$ \\
   & $\beta_1$ & $0.5851$ $(0.0902)^\text{a}$ & $0.5976$ $(0.0858)^\text{a}$ & $0.5596$ $(0.0898)^\text{a}$ \\
   & $\delta_{\text{50.0\%}}$ & $0.1469$ $(0.0552)^\text{a}$ & $0.1471$ $(0.0625)^\text{a}$ & $0.1484$ $(0.0600)^\text{a}$ \\
   & $\delta_{\text{63.5\%}}$ & $0.1398$ $(0.0553)^\text{a}$ & $0.1407$ $(0.0539)^\text{a}$ & $0.1374$ $(0.0511)^\text{a}$ \\
   & $\beta_2$ & $-0.2683$ $(0.0434)^\text{a}$ & $-0.2751$ $(0.0387)^\text{a}$ & $-0.2545$ $(0.0397)^\text{a}$ \\
 Dispersion &  &  &  & \\
   & $\alpha_{\text{37.5\%}}$ & $-0.0826$ $(0.1652)^{\phantom{\text{a}}}$ & $-0.3546$ $(0.3145)^{\phantom{\text{a}}}$ & $-0.7023$ $(0.4067)^{\phantom{\text{a}}}$ \\
   & $\alpha_{\text{50.0\%}}$ & -- & $-0.2979$ $(0.2982)^{\phantom{\text{a}}}$ & $-0.6413$ $(0.3979)^{\phantom{\text{a}}}$ \\
   & $\alpha_{\text{63.5\%}}$ & -- & $0.6893$ $(0.3023)^\text{a}$ & $0.4615$ $(0.3459)^{\phantom{\text{a}}}$ \\
   & $\gamma_1$ & -- & -- & $0.4314$ $(0.2983)^{\phantom{\text{a}}}$ \\
   \bottomrule
\end{tabularx}
\endgroup
\footnotesize \raggedright
 Est (SE)$^\text{a}$ indicates $|$Est$/$SE$|$ $> 1,96$.
\end{table}

The 95\% confidence intervals for the dispersion components are shown in
Figure \@ref(fig:profile-plot-soybean). The intervals are obtained by
working with the profile deviance function with respect to each
$\alpha_j$. The results indicate neither under- nor overdispersion for
moisture levels $37.5$\% and $50$\%. For the level $63.5$\%, there is
evidence for underdispersion, although the lower bound is close to zero.

```{r profile-plot-soybean, results="asis", fig.width=8.5, fig.height=3, fig.cap="Profile deviation for each dispersion component and their respective 95\\% confidence intervals."}
knitr::include_graphics("chapter3-figure/profile-plot-soybean-1")
```

The observed and fitted counts for each moisture level with confidence
intervals are shown in Figure \@ref(fig:fitted-curves-soybean), along
with the optimum doses. The results show that potassium doses at
$104.21$, $132.63$ and $132.63$mg dm$^{−3}$ lead to the expected maximum
number of pods for the moisture level $37.5$\%, $50$\%, and $62.5$\%,
respectively.  In terms of the number of pods, there is evidence that
fertilization with potassium can compensate the water deficit in soybean
culture.

```{r fitted-curves-soybean, results="asis", fig.width=8.5, fig.height=3.8, fig.cap="Fitted curves for the number of pods with 95\\% confidence intervals. The dotted lines indicate the optimal doses and the corresponding expected number of pods."}
knitr::include_graphics("chapter3-figure/fitted-curves-soybean-1")
```

In this analysis, we considered a quadratic predictor for the number of
pods along the potassium doses under which we could find the optimal
doses for each moisture level. However, in practice, this is not a
reasonable behavior as the number of pods decreases with increasing
doses of potassium. Here, a nonlinear predictor with a plateau or
asymptote can be very useful with direct interpretations, but is was not
explored here.

## Final remarks ## {#chap3-final-remarks}

In this paper, COM-Poisson models with varying dispersion were proposed
for jointly modeling mean and dispersion in the analysis of counts with
different levels of dispersion. This class of models allows modeling the
dispersion depending on covariates. The parameters are estimated by the
maximum likelihood method and inferences are made based on the
asymptotic distributions of the estimators.

We carried out a simulation study to assess the properties of the
maximum likelihood estimators obtained by two proposed strategies. The
results of our simulation study suggested that the maximum likelihood
estimators for the mean and dispersion regression parameters are unbiased
and consistent. The comparison of the strategies for maximum likelihood
estimation indicate that using the strategy (a) (maximization of
conditional likelihood) reduces the fitting times and leads to
inferences quite similar to the strategy (b) (maximization of full
likelihood). However, studies regarding the coverage rate of confidence
intervals and the evaluation of type 1 errors in hypothesis testing are
necessary.

The methodology is applied to analyze two count datasets obtained from
planned experiments. We compared the COM-Poisson models with double
generalized linear models and GAMLSS approach. The proposal presented
improvements in terms of fitting data when compared to the conventional
COM-Poisson with constant dispersion and was competitive with the double
generalized linear models and GALMSS based on double Poisson
distribution.

For future work, simulation studies are necessary in order to evaluate
the robustness of the model. In addition, as an improvement in the case
study analysis, considering nonlinear predictors for the mean can be
useful in order to avoid the use of high-order polynomials and to impose
biological restrictions.

\section*{References}
\vspace{-1cm}
\printbibliography[segment=\therefsegment,heading=subbibliography,title={\numberline{}References}]
