# Motivating studies # {#chapter0-data}

In this chapter, we present a number of data sets to illustrate some
scientific and statistical issues which arise from count data. These
data sets will be used throughout the text for illustration of the
proposed methodology.

```{r chapter0}
knitr::read_chunk("script-chapter0.R")
```

```{r chap0-packages}
```

## Artificial defoliation in cotton phenology ## {#chap0-cotton}

Cotton production can be drastically reduced by attack of defoliating
insects. Depending on the growth stage, the plants can recover from the
caused damage and keep production not affected or can have the
production reduced by low intensity defoliation. In order to study the
recovery of cotton plants (_Gossypium hirsutum_) in terms of production,
@Silva2012 conducted a greenhouse experiment under a completely
randomized design with five replicates. The experimental unity was a pot
with two plants and it was recorded the number of cotton bolls at five
artificial defoliation levels (0%, 25%, 50%, 75%, and 100%) and five
growth stages: vegetative, flower-bud, blossom, fig and cotton boll.

```{r chap0-plot-cotton, fig.width=9, fig.height=4, fig.cap="(a) Number of bolls produced for each artificial defoliation level and each growth stage (solid lines represent loess curves) and (b) sample variance against the sample mean of the five replicates for each combination of defoliation level and growth stage (dotted line is the identity line and solid line is the least square line)."}
```

Figure \@ref(fig:chap0-plot-cotton)(a) shows the number of cotton bolls
recorded for each combination of defoliation level and growth stage; the
smoothers indicate that are different quadratic effects by growth
stage. Figure \@ref(fig:chap0-plot-cotton)(b) show that all sample
variances are smaller than the sample means, suggesting underdispersion.

Alternatives analysis of this data have been proposed in the
literature. @Zeviani2014 analyzed it using the Gamma-Count
distribution. @Bonat2018 used this data for illustrating the extended
Poisson-Tweedie model. @Huang2017 and @RibeiroJr2018 analyzed it by
using different mean-parametrizations of the COM-Poisson model.

## Soil moisture and potassium doses on soybean culture ## {#chap0-soy}

In this second example we consider a study of potassium doses and soil
moisture levels on soybean (\emph{Glicine Max}) production. The tropical
soils are usually poor in potassium (K) and demand potassium
fertilization when cultivated with soybean to obtain satisfactory
yields. Soybean production is affected by long exposition to water
deficit. As potassium is a nutrient involved in the water balance in
plant, by hyphotesis, a good supply of potassium avoids to reduce
production.

To evaluate the effects of K doses and soil humidity levels on soybean
production, @Serafim2012 conducted a $5\times 3$ factorial experiment in
a randomized complete block design with 5 replicates. Five different
potassium doses (0, 0.3, 0.6, 1.2 and 1.8 $\times$ 100mg dm$^{-3}$) were
applied to the soil and soil moisture levels were controlled at (37.5,
50, and 62.5\%) . The experiment was carried out in a greenhouse and the
experimental units were pots with two plants in each. The count
responses measured were the total number of bean seeds per pot and the
number of viable pods.

```{r chap0-plot-soy, fig.width=8.5, fig.height=4.5, fig.cap="(a) Number of bean seeds and number of viable pods produced for each moisture level and each potassium dose (solid lines represent loess curves) and (b) sample variance against the sample mean of the five replicates for each combination of moisture level and potassium dose (dotted line is the identity line and solid line is the least-squares line)."}
```

Figure \@ref(fig:chap0-plot-soy)(a) shows the number of bean seeds and
the number of viable pods recorded for each combination of potassium
dose and moisture level; it is important to note the indication of a
quadratic effect of the potassium levels for both counts, as indicated
by the smoothers. For the number of bean seeds, most points in the
sample variance \textit{versus} sample means dispersion diagram
\@ref(fig:chap0-plot-soy)(b) are above the identity line, suggesting
overdispersion (block effect not yet removed). However, for the number
of viable pods, there are points above and below the identity line --
that leads to least-squares line very similar to the identity
line, suggesting that the equidispersion assumption is suitable for this
count.

## Assessing toxicity of nitrofen in aquatic systems ## {#chap0-nitrofen}

Nitrofen is a herbicide that was used extensively for the control of
broad-leaved and grass weeds in cereals and rice. Although it is
relatively non-toxic to adult mammals, nitrofen is a significant
tetragen and mutagen. It is also acutely toxic and reproductively toxic
to cladoceran zooplankton. Nitrofen is no longer in commercial use in
the U.S., having been the first pesticide to be withdrawn due to
tetragenic effects [@Bailer1994].

This data set comes from an experiment to measure the reproductive
toxicity of the herbicide, nitrofen, on a species of zooplankton
(_Ceriodaphnia dubia_). Fifty animals were randomized into
batches of ten and each batch was put in a solution with a measured
concentration of nitrofen ($0, 0.8, 1.6, 2.35$ and $3.10$
$\mu$g$/10^2$litre) (\texttt{dose}). Subsequently, the number of live
offspring was recorded.

```{r chap0-plot-nitrofen, include=FALSE, fig.width=9, fig.height=4, out.width="95%", fig.cap="(a) Number of live offspring observed for each nitrofen concentration level (solid lines represent loess curves) and (b) sample variance against sample mean of the 10 replicates for each concentration level (dotted line is the identity line, solid line is the least-squares line and the values between parentheses are the concentration levels)."}
```

```{r chap0-table-nitrofen, results="asis"}
```

Table \@ref(tab:chap0-table-nitrofen) shows the sample mean, sample
variance and sample dispersion index of the number of live offspring
obtained from the batches with different nitrofen concentration
level. This descriptive analysis indicates the nitrofen reduce the
number of live offsprings however it seems like the dispersion is also
influenced by the nitrofen concentration level (doses up to
$1.6\mu$g$/10^2$litre suggesting underdispersion while doses between
$1.6$ and $3.1\mu$g$/10^2$litre suggesting overdispersion). The Figure
\@ref(fig:chap0-plot-nitrofen) shows a graphical representation of these
results.

```{r chap0-plot-nitrofen, fig.width=8, fig.height=3.5, out.width="90%", fig.cap="(a) Number of live offsprings observed for each nitrofen concentration level and (b) scatterplot of the sample means against sample variances (dotted line is the identity line and solid line is the least-squares line)."}
```

## Evaluation of \textit{Annona mucosa} for control of \textit{Sitophilus zeamaus} ## {#chap0-sitophilus}

New control methods are necessary for stored grain pest management
programs due to both the widespread problems of insecticide-resistance
populations and the increasing concerns of consumers regarding pesticide
residues in food products. @Ribeiro2013 carried out an experiment to
assess the bioactivity of extracts of _Annona mucosa_ (Annonaceae) for
control _Sitophilus zeamaus_ (Coleoptera: Curculionidae), a major pest
of stored maize in Brazil. Petri dishes containing 10g of corn were
treated with extracts prepared with different parts of _Annona mucosa_
(seeds, leaves and branches) or just water (control) were completely
randomized with 10 replicates. Then 20 _Sitophilus zeamaus_ adults were
placed in each Petri dish and the numbers of emerged insects (progeny)
after 60 days were recorded.

```{r chap0-plot-sitophilus, include=FALSE, fig.width=9, fig.height=4.5, out.width="90%"}
```

```{r chap0-table-sitophilus, results="asis"}
```

For all treatments, the sample variance of the number of emerged insects
treatment is greater than their respective sample average, a strong
indication of overdispersion. For leaf extract, the sample variance is
three times higher than the mean.

This data set is presented and analysed by @Ribeiro2013 and later used
by @Demetrio2014 for illustrating the quasi-Poisson approach for
modelling overdispersed count data.

## Alternative substrats for bromeliad production ## {#chap0-bromelia}

Xaxim is a substrate used in the cultivation of bromeliads and orchids,
whose commercialization was prohibited in 2001. Since then, there has
been researching in botany to propose alternative substrates to Xaxim in
the cultivation of bromeliads, orchids, and other epiphytes
[@Salvador2008]. This dataset comes from a randomized experiment
conducted in a greenhouse in four blocks design with objective of
evaluate five different recipients of alternative substrates for
bromeliads [@Kanashiro2008]. All treatments contained peat and perlite
and differed in the third component: _Pinus_ bark, _Eucalyptus_ bark,
Coxim, coconut fiber and Xaxim. The variable of interest was the
number of leaves per experimental unit (pot with initially eight
plants), which was registered at 4, 173, 229, 285, 341, and 435 days
after planting.

```{r chap0-plot-bromelia, fig.width=8.7, fig.height=4, fig.cap="(a) Number of leaves for each treatment and day after planting (solid lines represent loess curves) and (b) sample variance against the sample mean per block (dotted line is the identity line and solid line is the least square line). "}
```

The observed number of leaves per each combination of days after
planting and treatments are shown in Figure
\@ref(fig:chap0-plot-bromelia)(a). There is a clear nonlinear
(sigmoidal) relationship between counts and days, mainly due to the low
number of leaves observed at 4 days (as expected). Figure
\@ref(fig:chap0-plot-bromelia)(b) shows the sample mean and sample
variances per block on the logarithm scale. All sample variances are
much smaller than their sample means, even on the logarithm scale,
indicating a strong evidence of underdispersion.

<!-- \section*{References} -->
<!-- \vspace{-1cm} -->
\printbibliography[segment=\therefsegment,heading=subbibliography,title={\numberline{}References}]
